\vspace{-0.0em} Move Schedules: Fast persistence \\ computations in coarse dynamic settings\thanks{This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}\vspace{-0.5em}

### 2023-10-14 src/move_schedules.tex
Reorganizing

%\documentclass{article} % for arxiv only
%\usepackage{arxiv} % for arxiv only 
%\pagestyle{plain} % arxiv only 
%%\usepackage[margin=1in]{geometry} % arxiv only

\PassOptionsToPackage{noend}{algorithmic}
\PassOptionsToPackage{noend}{algpseudocode}
\documentclass[sn-mathphys]{sn-jnl}

@@

%\usepackage[noend]{algorithmic}

% --- arxiv only ---- 
%\usepackage{xcolor}
%\usepackage{algorithm}% http://ctan.org/pkg/algorithms
%\PassOptionsToPackage{noend}{algpseudocode}
%\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx

@@

% --- Springer only ---
\title[Fast Persistence Computations in Dynamic Settings]{
	\vspace{-2.0em} Move Schedules: Fast persistence computations in coarse dynamic settings
	\footnote{
	\textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.
	}
}

% --- Arxiv only --- 
%\title{\vspace{-0.0em} Move Schedules: Fast persistence \\ computations  in coarse dynamic settings\thanks{This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}\vspace{-0.5em}}
%\author{Matt Piekenbrock\thanks{Khoury College of Computer Sciences, Northeastern University.}\;\; and Jose A. Perea$^\dagger$\thanks{Department of Mathematics and Khoury College of Computer Sciences, Northeastern University.}}
%\date{}

% --- Springer only ---
\author[1]{\fnm{Matthew} \sur{Piekenbrock}}\email{piekenbrock.m@northeastern.edu}
\affil[1]{
	\orgdiv{Khoury College of Computer Sciences}, 
	\orgname{Northeastern University}}

% --- Springer only --- 
\author[2]{\fnm{Jose} \sur{A. Perea}}\email{j.pereabenitez@northeastern.edu}
\affil[2]{
	\orgdiv{Department of Mathematics and Khoury College of Computer Sciences},
	\orgname{Northeastern University}}
% \orgaddress{\street{360 Huntington Ave}, \city{Boston}, \postcode{02115}, \state{MA}}
%}

@@

%\maketitle  % arxiv only, ensures abstract follows title

% REQUIRED - "Abstracts must be able to stand alone and so cannot contain citations to the paper's references, equations, etc. .... not exceeding 250 words"
%\begin{abstract} % arxiv only 
\abstract{ % springer only
Matrix reduction is the standard procedure for computing the persistent homology of a filtered simplicial complex with $m$ simplices. Its output is a particular decomposition of the total boundary matrix, from which the persistence diagrams and generating cycles are derived. 
	Persistence diagrams are known to vary continuously with respect to their input, motivating the  study of their computation for time-varying filtered complexes. Computing persistence dynamically can be reduced to maintaining a valid decomposition under adjacent transpositions in the filtration order. 
	Since there are $O(m^2)$ such transpositions, this maintenance procedure exhibits limited scalability and is often too fine for many applications. 
We propose a coarser strategy for maintaining the decomposition over a 1-parameter family of filtrations. By reduction to a particular longest common subsequence problem, we show that the minimal number of decomposition updates $d$ can be found in $O(m \log \log m)$ time and $O(m)$ space, and that the corresponding sequence of permutations---which we call a \emph{schedule}---can be constructed in $O(d m \log m)$ time.  
We also show that, in expectation, the storage needed to employ this strategy is actually sublinear in $m$. 
Exploiting this connection, we show experimentally that the decrease in operations to compute diagrams across a family of filtrations is proportional to the difference between the expected quadratic number of states and the proposed sublinear coarsening.
Applications to video data, dynamic metric space data, and multiparameter persistence are also presented.
} % springer only
%\end{abstract} % arxiv only

\maketitle %  Springer only, ensures abstract follows title

@@

% --- Springer only (REQUIRED) ---
\pacs[MSC Classification]{68T09, 55N31, 62R40}

@@

\begin{figure*}[!htb]
 	\centering
 	\includegraphics[width=0.75\textwidth]{circle_vineyards.png}
 	\caption{The cumulative column operations needed to compute persistence across the time-varying filtration of grayscale images. Observe 10 independent persistence computations evenly spaced in time (green line) captures the major topological changes and is the most computationally efficient approach shown.}
 	\label{fig:vineyards}
 \end{figure*}

@@

 The time complexity of \emph{vineyards} is determined entirely by the complexity of performing a single adjacent transposition.
Formally, since column operations are the largest complexity operations needed and each column can have potentially $O(m)$ entries, the complexity of \emph{vineyards} is $O(m)$ per transposition. 
Inspection of the individual cases of the algorithm from~\cite{cohen2006vines} shows that any single transposition requires at most two such operations on both $R$ and $V$.
However, several factors can affect the runtime efficiency of the vineyards algorithm. 
On the positive side, as both the matrices $R$ and $V$ are often sparse, the cost of a given column operation is proportional to the number of non-zero  entries in the two columns being modified. Moreover, as a rule of thumb, it has been observed that most transpositions require no column operations~\cite{edelsbrunner2000topological}. 
On the negative side, one needs to frequently query the non-zero status of various entries in $R$ and $V$ (consider evaluating e.g. Case 1.1 in~\cite{cohen2006vines}), which accrues a non-trivial runtime cost due to the quadratic frequency with which they are required.   
% Moreover, an efficient implementation of the vineyards algorithm requires a non-standard sparse matrix representation that allows swapping any two rows and columns in $O(1)$, as described in Figure 5 in~\cite{cohen2006vines}. 
% Though these minor operations do not affect the asymptotic complexity of the vineyards algorithm, they collectively accrue a non-trivial runtime cost

@@

For completeness, we recapitulate the motivation of the \emph{moves} algorithm from~\cite{busaryev2010tracking}. 
Let $f: K \rightarrow [m]$ denote a filtration of size $m = \lvert K \rvert$ and $R = DV$ its decomposition. 
%For any given matrix $A$, we denote its $i$th row (column) with $\mathrm{row}_A(i)$ ($\mathrm{col}_A(i)$, respectively).
Consider the permutation $P$ that moves a simplex $\sigma_i$ in $K$ to position $j$, shifting all intermediate simplices $\sigma_{i+1}, \dots, \sigma_{j}$ down by one ($i < j$). 
To perform this shift, all entries $V_{ik} \neq 0$ with column positions $k \in [i+1,j]$ need to set to zero, otherwise $PV$ is not upper-triangular. 
We may zero these entries in $V$ using column operations $V(\ast)$, ensuring invariant I2~\eqref{inv:decomposition} is maintained, however these operations may render $R' = P R(\ast) P^T$ unreduced, breaking invariant I3.
Of course, we could then reduce $R'$ with additional column operations, but the number of such operations scales $O(\lvert i - j\rvert^2)$, which no more efficient than simply performing the permutation and applying the reduction algorithm to columns $[i,j]$ in $R$.

@@

As shown by example 2.3, \emph{moves} can be cheaper than \emph{vineyards} in terms of column operations. However, it is not clear that this is always the case upon inspection of~\ref{alg:mr}, as the usage of a donor column seemingly implies that many $O(m)$ copy operations need to be performed. 
It turns out that we may handle all such operations except the first in $O(1)$ time, which we formalize below.
\begin{proposition}\label{prop:factor2}
	Let $(K,f)$ denote a filtration of size $\lvert K \rvert = m$ with decomposition $R = D V$ and let $T$ denote the number of column operations needed by \emph{vineyards} to perform the sequence of transpositions:
	$$ R = R_1 \overset{s_1}{\to} R_2 \overset{s_2}{\to} \dots \overset{s_k}{\to} R_{k-1} = R' $$
	where $s_i$ denotes the transposition $(i,i+1)$, $i < j$, and $k = \lvert i - j\rvert$. Moreover, let $M$ denote the number of column operations to perform the same update $R \mapsto R'$ with $\mathrm{Move}(i,j)$. Then the inequality $M \leq T$ holds.
\end{proposition}
\begin{proof} \normalsize
First, consider executing the \emph{vineyards} algorithm with a given pair $(i, j)$. As there are at most $2$ column operations, any contiguous sequence of transpositions $(i,i+1), (i+1,i+2), \ldots (j-1,j)$ induces at most $2(\lvert i - j \rvert)$ column operations in both $R$ and $V$, giving a total of $4(\lvert i - j \rvert)$ column operations.

Now consider a single MoveRight($i$,$j$) outlined in Algorithm~\ref{alg:mr}. Here, the dominant cost again are the column operations (line 5). 
Though we need an extra $O(m)$ storage allocation for the donor columns $d_\ast$ prior to the movement, notice that assignment to and from $d_\ast$ (lines (4), and (7) in \textit{RestoreLeft} and \textit{MoveRight}, respectively) requires just $O(1)$ time via a pointer swapping argument. That is, when $d_\mathrm{low}' < d_\mathrm{low}$, instead of copying $\mathrm{col}_\ast(k)$ to $d_\ast'$---which takes $O(m)$ time---we instead swap their column pointers in $O(1)$   prior to column operations. After the movement, $d_\ast$ contains the newly modified column and $\mathrm{col}_\ast(k)$ contains the unmodified donor $d_\ast'$, so the final donor swap also requires $O(1)$ time. 
Since at most one $O(m)$ column operation is required for each index in $[i, j]$, moving a column from $i$ to $j$ where $i < j$ requires at most $2(\lvert i - j \rvert)$ column operations for both $R$ and $V$. The claimed inequality follows. 
%since although permutations take $O(\lvert i - j\rvert)$ time we require only $1$ such permutation per movement. 
	%Identifying non-zero entries in the matrix occurs once, which takes $O(\lvert i - j \rvert \log(m))$ time.
\end{proof}
As a final remark, we note that the combination of \emph{MoveRight} and \emph{MoveLeft} enable efficient simplex additions or deletions to the underlying complex. In particular, given $K$ and a decomposition $R = DV$, obtaining a valid decomposition $R' = D' V'$ of $K' = K \cup \{\sigma\}$ can be achieved by appending its requisite elementary chains to $D$ and $V$, reducing them, and then executing \emph{MoveLeft}$(m+1, i)$ with $i = f'(\sigma)$. Dually, deleting a simplex $\sigma_i$ may be achieved via \emph{MoveRight} by moving $i$-th to the end of the decomposition and dropping the corresponding columns. 
% with a logarithmic loss in efficiency, although executing the movement has a higher constant factor. Nonetheless, we can still move a simplex arbitrarily within the given filtration with such a procedure,  which when paired with the augmented sparse matrix structure from~\cite{} implies one can move simplices within the filtration with just O(|i - j|) time permutation, as opposed the |i - j| separate O(1) permutations. Moreover, the ability to move a simplex earlier in the filtration enables a key preprocessing step which we introduce next which, as we shall prove, may greatly reduce the number of times the decomposition need be maintained.

@@

\subsection{Continuous setting}\label{sec:continuous_setting}
In the \emph{vineyards} setting, a given homotopy  $F : K \times [0,1] \to \mathbb{R}$ continuously interpolating between $(K, f)$ and $(K, f')$ is discretized into a set of critical events that alter the filtration order. 
As $F$ determines the number of distinct filtrations encountered during the deformation from $f$ to $f'$, a natural question is whether such an interpolation can be modified so as to minimize $\lvert \mathcal{S}\rvert$---the number of times the decomposition is restored to a valid state.
Towards explaining the phenomenon exhibited in Figure~\ref{fig:vineyards}, we begin by analyzing a particular class of interpolation schemes in order to establish an upper bound on this quantity.

@@

If our goal is to decrease  $\lvert S_F \rvert$, one option is to \emph{coarsen} $S_F$ to a new schedule $\widetilde{S}_F$ by e.g. collapsing contiguous sequences of adjacent transpositions to moves, via the map:
\begin{equation}\label{eq:tr_to_mv}
	 (i, i+1)(i+1, i+2)\cdots(j-1, j) \mapsto (j, i+1, \cdots, j-1, i)  \quad \text{ if } i < j
\end{equation}
Clearly $\lvert \widetilde{S}_F \rvert \leq \lvert S_F \rvert$ and the associated coarsened $\widetilde{S}_F$ requires just $O(m)$ time to compute. However, the coarsening depends entirely on the initial choice of $F$ and the quadratic upper bound remains---it is always possible that there are no contiguous subsequences to collapse. 
This suggests one must either abandon the continuous  setting or make stronger assumptions on $F$ to have any hope of keeping $\lvert S_F \rvert \sim O(m)$ in size.
% Indeed, although many transpositions require no column reductions in practice, detecting the existence of non-zero entries at specific positions in the sparse matrix takes $O(\log(m))$ time per transposition.

@@

In cycle notation, this corresponds to the cyclic permutation:
\begin{equation}\label{eq:move_permutation}
m_{ij} = (\enspace i \enspace j \enspace j\text{-}1 \enspace \dots \enspace i\text{+}2 \enspace i\text{+}1 \enspace )

@@

First, we require a few definitions. Recall that a \emph{sorting} $\mathcal{S}$ with respect to two permutations $p, q \in S_m$ is an ordered sequence of permutations $\mathcal{S} = (s_1, s_2, \dots, s_d)$ satisfying $q = s_d \circ \dots s_1 \circ p$.
By definition, a subsequence in $\mathcal{L}$ common to both $p$ and $q$ satisfies:
\begin{equation}
	p^{-1}(\sigma) < p^{-1}(\tau) \implies q^{-1}(\sigma) < q^{-1}(\tau) \quad \forall \sigma, \tau \in \mathcal{L}
\end{equation}
 where $p^{-1}(\sigma)$ (resp. $q^{-1}(\sigma)$) denotes the position of $\sigma$ in $p$ (resp. $q$). Thus, obtaining a sorting $p \mapsto q$ of size $d = m - \lvert \mathcal{L} \rvert$ reduces to applying a sequence of moves in the complement of $\mathcal{L}$. 
 Formally, we define a permutation $s \in S_m$ as a \emph{valid} operation with respect to a fixed pair $p, q \in S_m$ if:
 \begin{equation}\label{eq:valid_move}
 	\lvert \mathrm{LCS}(s \circ p, q) \rvert = \lvert \mathrm{LCS}(p,q) \rvert + 1
 \end{equation}
  The problem of constructing a sorting $\mathcal{S}$ of size $d$ thus reduces to the problem of choosing a sequence of $d$ valid moves, which we call a \emph{valid sorting}. 
  To do this efficiently, let $\mathcal{U}$ denote an ordered set-like data structure that supports the following operations on elements $\sigma \in M$ from the set $M = \{0, 1, \dots, m+1\}$:
  \begin{enumerate}
  	\item $\mathcal{U} \cup \sigma$---inserts $\sigma$ into $\mathcal{U}$,
  	\item $\mathcal{U} \setminus \sigma$---removes $\sigma$ from $\mathcal{U}$,
  	\item $\mathcal{U}_{\mathrm{succ}}(\sigma)$---obtain the successor of $\sigma$ in $\mathcal{U}$, if it exists, otherwise return $m+1$
  	\item $\mathcal{U}_{\mathrm{pred}}(\sigma)$---obtain the predecessor of $\sigma$ in $\mathcal{U}$, if it exists, otherwise return $0$
  \end{enumerate}
  Given $\mathcal{U}$, a valid sorting can be constructed by querying and maintaining information about the $\mathrm{LCS}$ in $\mathcal{U}$. To see this, suppose $\mathcal{U}$ contains the current $\mathrm{LCS}$ between two permutations $p$ and $q$. By definition of the LCS, we have:
 \begin{equation}\label{eq:valid_move_ineq}
 	p^{-1}(\mathcal{U}_{\mathrm{pred}}(\sigma)) < p^{-1}(\sigma) < p^{-1}(\mathcal{U}_{\mathrm{succ}}(\sigma))
 \end{equation}
  for every $\sigma \in \mathcal{U}$. Now, suppose we choose some element $\sigma \notin \mathcal{U}$ which we would like to add to the LCS. If $p^{-1}(\sigma) < p^{-1}(\mathcal{U}_{\mathrm{pred}}(\sigma))$, then we must move $\sigma$ to the right of its predecessor in $p$ such that~\eqref{eq:valid_move_ineq} holds. Similarly, if  $p^{-1}(\mathcal{U}_{\mathrm{succ}}(\sigma)) < p^{-1}(\sigma)$, then we must move $\sigma$ left of its successor in $p$. Assuming the structure $\mathcal{U}$ supports all of the above operations in $O(\log m)$ time, we easily deduce a $O(d m\log m)$ algorithm for obtaining a valid sorting.
%  which for completeness is shown via Algorithm~\ref{alg:sorting} in the appendix. 
  % one can incrementally build a valid sorting of size $d$ by arbitrarily choosing valid move operations . all elements in $\mathrm{LCS}(p,q)$ into $\mathcal{T}$
%  incrementally updating the inverse permutation $p^{-1}$ of $p$. Note that   
 % Further, let $\mathcal{I}_p$ denote an array of length $m = \lvert p \rvert$ whose values $\mathcal{I}(\sigma)$ store the position of the element $\sigma$ in $p$. Assume also that $\mathcal{I}_p()$  
 %for every pair of elements $\sigma, \tau \in \mathcal{L}$, if $p_{\sigma} < p_{\tau}$ where $p_{\sigma}$ denotes the the position of $\sigma$ in $p$, then $q_{\sigma} < q_{\tau}$ must also be true. 
 %under certain insertion and deletion operations.

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=0.7\textwidth, height=12.15em]{boid_sim_results.png} % arxiv only
%	\includegraphics[width=\textwidth, height=16.15em]{boid_sim_results.png}
	\caption{On the left, the cumulative number of column operations (log-scale) of the three baseline approaches tested. On the right, the normalized $K_\tau$ between adjacent filtrations depicts the coarseness of the discretization---about $5\%$ of the $\approx O(m^2)$ simplex pairs between adjacent filtrations are discordant.}
	\label{fig:boid_sim_results}
\end{figure}

>>>>

\documentclass{article} % for arxiv only
\usepackage{arxiv} % for arxiv only 
\pagestyle{plain} % arxiv only 
%\usepackage[margin=1in]{geometry} % arxiv only

%\PassOptionsToPackage{noend}{algorithmic} % springer only 
%\PassOptionsToPackage{noend}{algpseudocode} % springer only 
%\documentclass[sn-mathphys]{sn-jnl} % springer only

@@

%\usepackage[noend]{algorithmic} % arxiv only

% --- arxiv only ---- 
\usepackage{xcolor}
\usepackage{algorithm}% http://ctan.org/pkg/algorithms
\PassOptionsToPackage{noend}{algpseudocode}
\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx

@@

% --- Springer only ---
%\title[Fast Persistence Computations in Dynamic Settings]{
%	\vspace{-2.0em} Move Schedules: Fast persistence computations in coarse dynamic settings
%	\footnote{
%	\textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.
%	}
%}

% --- Arxiv only --- 
\title{\vspace{-0.0em} Move Schedules: Fast persistence \\ computations  in coarse dynamic settings\thanks{This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}\vspace{-0.5em}}
\author{Matt Piekenbrock\thanks{Khoury College of Computer Sciences, Northeastern University.}\;\; and Jose A. Perea$^\dagger$\thanks{Department of Mathematics and Khoury College of Computer Sciences, Northeastern University.}}
\date{}

% --- Springer only ---
%\author[1]{\fnm{Matthew} \sur{Piekenbrock}}\email{piekenbrock.m@northeastern.edu}
%\affil[1]{
%	\orgdiv{Khoury College of Computer Sciences}, 
%	\orgname{Northeastern University}}

% --- Springer only --- 
%\author[2]{\fnm{Jose} \sur{A. Perea}}\email{j.pereabenitez@northeastern.edu}
%\affil[2]{
%	\orgdiv{Department of Mathematics and Khoury College of Computer Sciences},
%	\orgname{Northeastern University}}
% \orgaddress{\street{360 Huntington Ave}, \city{Boston}, \postcode{02115}, \state{MA}}
%}

@@

\maketitle  % arxiv only, ensures abstract follows title

% REQUIRED - "Abstracts must be able to stand alone and so cannot contain citations to the paper's references, equations, etc. .... not exceeding 250 words"
\begin{abstract} % arxiv only 
%\abstract{ % springer only
Matrix reduction is the standard procedure for computing the persistent homology of a filtered simplicial complex with $m$ simplices. Its output is a particular decomposition of the total boundary matrix, from which the persistence diagrams and generating cycles are derived. 
	Persistence diagrams are known to vary continuously with respect to their input, motivating the  study of their computation for time-varying filtered complexes. Computing persistence dynamically can be reduced to maintaining a valid decomposition under adjacent transpositions in the filtration order. 
	Since there are $O(m^2)$ such transpositions, this maintenance procedure exhibits limited scalability and is often too fine for many applications. 
We propose a coarser strategy for maintaining the decomposition over a 1-parameter family of filtrations. By reduction to a particular longest common subsequence problem, we show that the minimal number of decomposition updates $d$ can be found in $O(m \log \log m)$ time and $O(m)$ space, and that the corresponding sequence of permutations---which we call a \emph{schedule}---can be constructed in $O(d m \log m)$ time.  
We also show that, in expectation, the storage needed to employ this strategy is actually sublinear in $m$. 
Exploiting this connection, we show experimentally that the decrease in operations to compute diagrams across a family of filtrations is proportional to the difference between the expected quadratic number of states and the proposed sublinear coarsening.
Applications to video data, dynamic metric space data, and multiparameter persistence are also presented.
%} % springer only
\end{abstract} % arxiv only

%\maketitle %  Springer only, ensures abstract follows title

@@

% --- Springer only (REQUIRED) ---
%\pacs[MSC Classification]{68T09, 55N31, 62R40}

@@

\begin{figure*}[!htb]
 	\centering
% 	\includegraphics[width=0.75\textwidth]{circle_vineyards.png} % springer only 
	\includegraphics[width=0.70\textwidth]{circle_vineyards.png}
 	\caption{The cumulative column operations needed to compute persistence across the time-varying filtration of grayscale images. Observe 10 independent persistence computations evenly spaced in time (green line) captures the major topological changes and is the most computationally efficient approach shown.}
 	\label{fig:vineyards}
 \end{figure*}

@@

 The time complexity of \emph{vineyards} is determined entirely by the complexity of performing  adjacent transpositions.
Since column operations are the largest complexity operations needed and each column can have potentially $O(m)$ entries, the complexity of \emph{vineyards} is $O(m)$ per transposition. 
Inspection of the individual cases of the algorithm from~\cite{cohen2006vines} shows that any single transposition requires at most two such operations on both $R$ and $V$.
However, several factors can affect the runtime efficiency of the vineyards algorithm. 
On the positive side, as both the matrices $R$ and $V$ are often sparse, the cost of a given column operation is proportional to the number of non-zero  entries in the two columns being modified. Moreover, as a rule of thumb, it has been observed that most transpositions require no column operations~\cite{edelsbrunner2000topological}. 
On the negative side, one needs to frequently query the non-zero status of various entries in $R$ and $V$ (consider evaluating e.g. Case 1.1 in~\cite{cohen2006vines}), which accrues a non-trivial runtime cost due to the quadratic frequency with which they are required.   
% Moreover, an efficient implementation of the vineyards algorithm requires a non-standard sparse matrix representation that allows swapping any two rows and columns in $O(1)$, as described in Figure 5 in~\cite{cohen2006vines}. 
% Though these minor operations do not affect the asymptotic complexity of the vineyards algorithm, they collectively accrue a non-trivial runtime cost

@@

For completeness, we recapitulate the motivation of the \emph{moves} algorithm from~\cite{busaryev2010tracking}. 
Let $f: K \rightarrow [m]$ denote a filtration of size $m = \lvert K \rvert$ and $R = DV$ its decomposition. 
%For any given matrix $A$, we denote its $i$th row (column) with $\mathrm{row}_A(i)$ ($\mathrm{col}_A(i)$, respectively).
Consider the permutation $P$ that moves a simplex $\sigma_i$ in $K$ to position $j$, shifting all intermediate simplices $\sigma_{i+1}, \dots, \sigma_{j}$ down by one ($i < j$). 
To perform this shift, all entries $V_{ik} \neq 0$ with column positions $k \in [i+1,j]$ need to set to zero, otherwise $PV$ is not upper-triangular. 
We may zero these entries in $V$ using column operations $V(\ast)$, ensuring invariant I2~\eqref{inv:decomposition} is maintained, however these operations may render $R' = P R(\ast) P^T$ unreduced, breaking invariant I3.
Of course, we could then reduce $R'$ with additional column operations, but the number of such operations scales $O(\lvert i - j\rvert^2)$, but this is no more efficient than simply performing the permutation and applying the reduction algorithm to columns $[i,j]$ in $R$.

@@

As shown by example 2.3, \emph{moves} can be cheaper than \emph{vineyards} in terms of column operations. However, it is not clear that this is always the case upon inspection of Algorithm~\ref{alg:mr}, as the usage of a donor column seemingly implies that many $O(m)$ copy operations need to be performed. 
It turns out that we may handle all such operations except the first in $O(1)$ time, which we formalize below.
\begin{proposition}\label{prop:factor2}
	Let $(K,f)$ denote a filtration of size $\lvert K \rvert = m$ with decomposition $R = D V$ and let $T$ denote the number of column operations needed by \emph{vineyards} to perform the sequence of transpositions:
	$$ R = R_1 \overset{s_1}{\to} R_2 \overset{s_2}{\to} \dots \overset{s_k}{\to} R_{k-1} = R' $$
	where $s_i$ denotes the transposition $(i,i+1)$, $i < j$, and $k = \lvert i - j\rvert$. Moreover, let $M$ denote the number of column operations to perform the same update $R \mapsto R'$ with $\mathrm{Move}(i,j)$. Then the inequality $M \leq T$ holds.
\end{proposition}
\begin{proof} \normalsize
First, consider executing the \emph{vineyards} algorithm with a given pair $(i, j)$. As there are at most $2$ column operations, any contiguous sequence of transpositions $(i,i+1), (i+1,i+2), \ldots (j-1,j)$ induces at most $2(\lvert i - j \rvert)$ column operations in both $R$ and $V$, giving a total of $4(\lvert i - j \rvert)$ column operations.

Now consider MoveRight($i$,$j$), outlined in Algorithm~\ref{alg:mr}. Here, the dominant cost again are the column operations (line 5). 
Though we need an extra $O(m)$ storage allocation for the donor columns $d_\ast$ prior to the movement, notice that assignment to and from $d_\ast$ (lines (4), and (7) in \textit{RestoreLeft} and \textit{MoveRight}, respectively) requires just $O(1)$ time via a pointer swapping argument. That is, when $d_\mathrm{low}' < d_\mathrm{low}$, instead of copying $\mathrm{col}_\ast(k)$ to $d_\ast'$---which takes $O(m)$ time---we instead swap their column pointers in $O(1)$   prior to column operations. After the movement, $d_\ast$ contains the newly modified column and $\mathrm{col}_\ast(k)$ contains the unmodified donor $d_\ast'$, so the final donor swap also requires $O(1)$ time. 
Since at most one $O(m)$ column operation is required for each index in $[i, j]$, moving a column from $i$ to $j$ where $i < j$ requires at most $2(\lvert i - j \rvert)$ column operations for both $R$ and $V$. The claimed inequality follows. 
%since although permutations take $O(\lvert i - j\rvert)$ time we require only $1$ such permutation per movement. 
	%Identifying non-zero entries in the matrix occurs once, which takes $O(\lvert i - j \rvert \log(m))$ time.
\end{proof}
As a final remark, we note that the combination of \emph{MoveRight} and \emph{MoveLeft} enable efficient simplex additions or deletions to the underlying complex. In particular, given $K$ and a decomposition $R = DV$, obtaining a valid decomposition $R' = D' V'$ of $K' = K \cup \{\sigma\}$ can be achieved by appending its requisite elementary chains to $D$ and $V$, reducing them, and then executing \emph{MoveLeft}$(m+1, i)$ with $i = f'(\sigma)$. Dually, deleting a simplex $\sigma_i$ may be achieved via \emph{MoveRight} by moving $i$-th to the end of the decomposition and dropping the corresponding columns. 
% with a logarithmic loss in efficiency, although executing the movement has a higher constant factor. Nonetheless, we can still move a simplex arbitrarily within the given filtration with such a procedure,  which when paired with the augmented sparse matrix structure from~\cite{} implies one can move simplices within the filtration with just O(|i - j|) time permutation, as opposed the |i - j| separate O(1) permutations. Moreover, the ability to move a simplex earlier in the filtration enables a key preprocessing step which we introduce next which, as we shall prove, may greatly reduce the number of times the decomposition need be maintained.

@@

\subsection{Continuous setting}\label{sec:continuous_setting}
In the \emph{vineyards} setting, a given homotopy  $F : K \times [0,1] \to \mathbb{R}$ continuously interpolating between $(K, f)$ and $(K, f')$ is discretized into a set of critical events that alter the filtration order. 
As $F$ determines the number of distinct filtrations encountered during the deformation from $f$ to $f'$, a natural question is whether such an interpolation can be modified so as to minimize $\lvert \mathcal{S}\rvert$---the number of times the decomposition is restored to a valid state.
Towards explaining the phenomenon exhibited in Figure~\ref{fig:vineyards}, we begin by analyzing a class of interpolation schemes to establish an upper bound on this quantity.

@@

If our goal is to decrease  $\lvert S_F \rvert$, one option is to \emph{coarsen} $S_F$ to a new schedule $\widetilde{S}_F$ by e.g. collapsing contiguous sequences of adjacent transpositions to moves, via the map $(i, i+1)(i+1, i+2)\cdots(j-1, j) \mapsto (j, i+1, \cdots, j-1, i)$ (if $i < j$).
%\end{equation}
Clearly $\lvert \widetilde{S}_F \rvert \leq \lvert S_F \rvert$ and the associated coarsened $\widetilde{S}_F$ requires just $O(m)$ time to compute. However, the coarsening depends entirely on the initial choice of $F$ and the quadratic upper bound remains---it is always possible that there are no contiguous subsequences to collapse. 
This suggests one must either abandon the continuous  setting or make stronger assumptions on $F$ to have any hope of keeping $\lvert S_F \rvert \sim O(m)$ in size.
% Indeed, although many transpositions require no column reductions in practice, detecting the existence of non-zero entries at specific positions in the sparse matrix takes $O(\log(m))$ time per transposition.

@@

In cycle notation, this corresponds to the cyclic permutation $m_{ij} = (\enspace i \enspace j \enspace j\text{-}1 \enspace \dots \enspace i\text{+}2 \enspace i\text{+}1 \enspace )$.
%\begin{equation}\label{eq:move_permutation}
%m_{ij} = (\enspace i \enspace j \enspace j\text{-}1 \enspace \dots \enspace i\text{+}2 \enspace i\text{+}1 \enspace )

@@

Recall that a \emph{sorting} $\mathcal{S}$ with respect to two permutations $p, q \in S_m$ is an ordered sequence of permutations $\mathcal{S} = (s_1, s_2, \dots, s_d)$ satisfying $q = s_d \circ \dots s_1 \circ p$.
By definition, a subsequence in $\mathcal{L}$ common to both $p$ and $q$ satisfies:
\begin{equation}
	p^{-1}(\sigma) < p^{-1}(\tau) \implies q^{-1}(\sigma) < q^{-1}(\tau) \quad \forall \sigma, \tau \in \mathcal{L}
\end{equation}
 where $p^{-1}(\sigma)$ (resp. $q^{-1}(\sigma)$) denotes the position of $\sigma$ in $p$ (resp. $q$). Thus, obtaining a sorting $p \mapsto q$ of size $d = m - \lvert \mathcal{L} \rvert$ reduces to applying a sequence of moves in the complement of $\mathcal{L}$. 
 Formally, we define a permutation $s \in S_m$ as a \emph{valid} operation with respect to a fixed pair $p, q \in S_m$ if it satisfies:
 \begin{equation}\label{eq:valid_move}
 	\lvert \mathrm{LCS}(s \circ p, q) \rvert = \lvert \mathrm{LCS}(p,q) \rvert + 1
 \end{equation}
  The problem of constructing a sorting $\mathcal{S}$ of size $d$ thus reduces to choosing a sequence of $d$ valid moves, which we call a \emph{valid sorting}. 
  To do this efficiently, let $\mathcal{U}$ denote an ordered set-like data structure that supports the following operations on elements $\sigma \in M$ from the set $M = \{0, 1, \dots, m+1\}$:
  \begin{enumerate}
  	\item $\mathcal{U} \cup \sigma$---inserts $\sigma$ into $\mathcal{U}$,
  	\item $\mathcal{U} \setminus \sigma$---removes $\sigma$ from $\mathcal{U}$,
  	\item $\mathcal{U}_{\mathrm{succ}}(\sigma)$---obtain the successor of $\sigma$ in $\mathcal{U}$, if it exists, otherwise return $m+1$
  	\item $\mathcal{U}_{\mathrm{pred}}(\sigma)$---obtain the predecessor of $\sigma$ in $\mathcal{U}$, if it exists, otherwise return $0$
  \end{enumerate}
  Given $\mathcal{U}$, a valid sorting can be constructed by querying and maintaining information about the $\mathrm{LCS}$ in $\mathcal{U}$. To see this, suppose $\mathcal{U}$ contains the current $\mathrm{LCS}$ between two permutations $p$ and $q$. By definition of the LCS, we have:
 \begin{equation}\label{eq:valid_move_ineq}
 	p^{-1}(\mathcal{U}_{\mathrm{pred}}(\sigma)) < p^{-1}(\sigma) < p^{-1}(\mathcal{U}_{\mathrm{succ}}(\sigma))
 \end{equation}
  for every $\sigma \in \mathcal{U}$. Now, suppose we choose some element $\sigma \notin \mathcal{U}$ which we would like to add to the LCS. If $p^{-1}(\sigma) < p^{-1}(\mathcal{U}_{\mathrm{pred}}(\sigma))$, then we must move $\sigma$ to the right of its predecessor in $p$ such that~\eqref{eq:valid_move_ineq} holds. Similarly, if  $p^{-1}(\mathcal{U}_{\mathrm{succ}}(\sigma)) < p^{-1}(\sigma)$, then we must move $\sigma$ left of its successor in $p$. Assuming the structure $\mathcal{U}$ supports all of the above operations in $O(\log m)$ time, we easily deduce a $O(d m\log m)$ algorithm for obtaining a valid sorting.
%  which for completeness is shown via Algorithm~\ref{alg:sorting} in the appendix. 
  % one can incrementally build a valid sorting of size $d$ by arbitrarily choosing valid move operations . all elements in $\mathrm{LCS}(p,q)$ into $\mathcal{T}$
%  incrementally updating the inverse permutation $p^{-1}$ of $p$. Note that   
 % Further, let $\mathcal{I}_p$ denote an array of length $m = \lvert p \rvert$ whose values $\mathcal{I}(\sigma)$ store the position of the element $\sigma$ in $p$. Assume also that $\mathcal{I}_p()$  
 %for every pair of elements $\sigma, \tau \in \mathcal{L}$, if $p_{\sigma} < p_{\tau}$ where $p_{\sigma}$ denotes the the position of $\sigma$ in $p$, then $q_{\sigma} < q_{\tau}$ must also be true. 
 %under certain insertion and deletion operations.

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=0.6\textwidth, height=14.15em]{boid_sim_results.png} % arxiv only
%	\includegraphics[width=\textwidth, height=16.15em]{boid_sim_results.png}
	\caption{On the left, the cumulative number of column operations (log-scale) of the three baseline approaches tested. On the right, the normalized $K_\tau$ between adjacent filtrations depicts the coarseness of the discretization---about $5\%$ of the $\approx O(m^2)$ simplex pairs between adjacent filtrations are discordant.}
	\label{fig:boid_sim_results}
\end{figure}
